---
title: "hazrd"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/amorris28/hazRd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/amorris28/hazRd/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`hazrd` is an R package for evaluating and visualizing Polygenic Hazard Score
(PHS) analyses. It provides a consistent, tidy-first API for computing
survival-based discrimination metrics, bootstrapped confidence intervals, and
Kaplan-Meier curves.

## Installation

### Latest release

Install the current stable release directly from GitHub:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("amorris28/hazrd")
```

Installing from the default (`main`) branch is equivalent to installing the
latest tagged release.

### Development version

To get the current development version (may contain unreleased features):

```{r, eval=FALSE}
devtools::install_github("amorris28/hazrd@0.2.1-dev")
```

## Getting started

For a full introduction see the vignette on the
[hazrd pkgdown website](https://amorris28.github.io/hazrd/).

## Quick example

```{r, eval=FALSE}
library(hazrd)

# Compute HR, C-index, OR, and HR_SD in one call
phs_metrics(
  test_data,
  metrics = c("HR", "C_index", "OR", "HR_SD"),
  or_age  = 70)

# Bootstrapped confidence intervals
phs_metrics(
  test_data,
  metrics   = c("HR", "C_index"),
  bootstrap = TRUE,
  n_boot    = 999,
  seed      = 42
)

# Kaplan-Meier plot stratified by PHS percentile (returns a ggplot object)
phs_km_curve(test_data)

# Return tidy data for a custom ggplot
km_data <- phs_km_curve(test_data, breaks = c(0.20, 0.40, 0.60, 0.80), output = "data")
```

## Planned features

The following are on the roadmap for future releases. Contributions welcome.

- **`phs_abs_risk()`** — absolute (cumulative incidence) risk curves with
  competing risk support (`km`, `aalen_johansen`, `fine_gray` methods)
- **`phs_percentile()`** — compute percentile ranks; optionally relative to a
  reference population via `ref_data`
- **`phs_cut()`** — assign individuals to percentile strata with human-readable
  factor labels
- **`phs_describe()`** — descriptive statistics (n, events, median follow-up,
  PHS mean/SD) by stratum
- **`phs_calibration()`** — calibration plot and Hosmer-Lemeshow statistic
- **`theme_phs()` / `scale_color_phs()`** — default ggplot2 theme and colour
  scale for hazrd plots
- **Uno’s C-index** — IPCW-weighted concordance for heavily censored data
  (`cindex_method = "uno"` in `phs_metrics()`)
- **Parallel bootstrap** — `parallel = "multicore"` / `"snow"` via
  `boot::boot()` (argument exists; only `"no"` is currently active)
- **Risk table** — numbers-at-risk appended below KM curves
  (`risk_table = TRUE` in `phs_km()`)
- **Alternative HR methods** — `continuous_point` and `categorical`
  (`hr_method` argument in `phs_metrics()`; stubs exist)
