---
title: "hazrd"
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/amorris28/hazRd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/amorris28/hazRd/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of hazrd is to simplify and standardize the development and testing of polygenic hazard score models in an opinionated way based mainly on past work in the [PHS repo](https://github.com/cmig-research-group/phs). Currently, most of the functionality focuses on external validation of PHS models. This includes calculation of standard statistics (80-20 hazard ratios, 80-20 odds ratios, and concordance index) and plotting of Kaplan-Meier curves and cumulative incidence curves. In the future, this will be expanded to model development, internal validation, imaging risk scores, multimodal hazard scores, and functions related to Digital Avatar.

## Installation

You can install the development version of hazrd from [GitHub](https://github.com/) by cloning the repository:

```{bash, eval=FALSE}
git clone git@github.com:amorris28/hazrd.git
```

and then opening `R` and installing the package:

```{r, eval=FALSE}
devtools::install("hazrd")
```

Replace `"hazrd"` with the path to the repository on your local computer.

## Getting Started

First, generate some test data. Here, we are using a simulated dataset that is included with the `hazrd` package.

```{r}
library(ggplot2)
library(hazrd)
set.seed(4649580)

head(test_data)
```

Next, plot the histogram of PHSes by case/control status.

```{r phs_hist}
phs_hist(test_data, normalize = TRUE)
```

Then, calculate the hazard ratio comparing the mean of the top 20% of PHSes to the mean of the bottom 20% (i.e., `HR80_20`). We can also generate 95% confidence intervals using bootstrapping.

```{r}
HR80_20 = get_hr(test_data, CI = TRUE, boot = 300)
str(HR80_20)
```

Each `get_` function also returns the output from each bootstrap iteration in `$iters` so that the user can plot these or calculate their own confidence intervals:

```{r HR_histogram}
ggplot(mapping = aes(x = HR80_20$iters)) +
    geom_histogram(binwidth = 2/3) +
    geom_vline(xintercept = HR80_20$HR, color = "red") +
    geom_vline(xintercept = HR80_20$conf.low) +
    geom_vline(xintercept = HR80_20$conf.high) +
        theme_minimal() +
        labs(x = "HR80_20", y = "Count")
```

Similarly, calculate the odds ratio at age 70 between the top 20% and bottom 20% of PHSes.

```{r}
OR80_20 = get_or(test_data, or_age = 70, CI = TRUE, boot = 300)
str(OR80_20)
```

```{r OR_histogram}
ggplot(mapping = aes(x = OR80_20$iters)) +
    geom_histogram(binwidth = 3/2) +
    geom_vline(xintercept = OR80_20$OR, color = "red") +
    geom_vline(xintercept = OR80_20$conf.low) +
    geom_vline(xintercept = OR80_20$conf.high) +
        theme_minimal() +
        labs(x = "OR80_20", y = "Count")
```

Return the concordance index with 95% confidence intervals from a coxph fit:

```{r}
c_index = get_cindex(test_data, CI = TRUE, boot = 300)
str(c_index)
```

Finally, plot the Kaplan-Meier curves with confidence intervals for centiles of interest.

```{r km_curve}
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
    x = x * 100
    y = y * 100
    out = paste0("PHS ", x, "-", y, "th centile")
    return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
    curven <- km_curve(data = test_data,  
                       interval = curves[[i]],
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
    curven$label = label_generator(curves[1, i], curves[2, i])
    km_curves = rbind(km_curves, curven)
}


ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
    geom_ribbon(alpha = 0.1,
                color = 0) +
    geom_step() +
    theme_minimal() +
    xlim(40, 100) + 
    ylim(0, 1) +
    labs(x = "Age", y = "Disease-free Survival") +
    scale_color_brewer(palette = "Set1",
                       name = "Centile") +
    scale_fill_brewer(palette = "Set1",
                       name = "Centile")
```

## Developer Instructions

These are general instructions for how to create and document an R package.

```{r, eval=FALSE}
install.packages(c("usethis", "devtools", "roxygen2"))
```

First, navigate to where you want to create the R package project directory. Open up an `R` console and run the `create_package` command. The first argument will be the name of the package and the name of the directory that is created within your current working directory.

```{r, eval=FALSE}
usethis::create_package("my_package")
```

Next, navigate to your package directory (`cd my_package`) and develop your package by adding code to the `R/` directory. Here is a simple example of how to structure an `.R` script within the `R/` directory called `my_function.R`.

```{r, eval=FALSE}
#' This is the Title of the Help Page for my_function
#'
#' This is a description of what this function does.
#'
#' @param x Description of what the "x" parameter expects, default value, whether it is optional
#' @param y Description of what the "y" parameter expects, etc.
#' @return Description of what this function returns
#' @examples
#' my_results <- my_function(x = 1, y = 2)
#' @export
my_function <- function(x, y) {
 # Some code
}
```

Once you have created some files in the `R/` directory, you can automatically generate documentation using either `roxygen2::roxygenise` or `devtools::document`. `document` is generally preferred and actually calls `roxygenise` as part of its testing. You can either call it without any arguments from the root directory of your R package or you can specify the path to the package as the first argument.

```{r, eval=FALSE}
devtools::document("path_to_your_package")
```

To verify there are no errors in the code or documentation, you can run `check`:

```{r, eval=FALSE}
devtools::check()
```

There is a [known bug](https://stackoverflow.com/questions/63613301/r-cmd-check-note-unable-to-verify-current-time) where `check` returns 1 Note that says "unable to verify current time". The workaround is to add `_R_CHECK_SYSTEM_CLOCK_=0` to your R environment file:

```{r, eval=FALSE}
usethis::edit_r_environ()
```

Resolve this and any other Errors, Warnings, and Notes that come up from `check`.

Finally, you can install the package locally using `devtools`:

```{r, eval=FALSE}
devtools::install("path_to_your_package")
```
