---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hazrd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hazrd)
library(ggplot2)
```

## Getting started with `{hazrd}`

Here is our test data included in the package.

```{r}
head(test_data)
```

Next, plot the histogram of PHSes by case/control status.

```{r phs_hist, fig.width=4}
phs_hist(test_data, normalize = TRUE)
```

### Generating test statistics and confidence intervals

Then, calculate the hazard ratio comparing the mean of the top 20% of PHSes to the mean of the bottom 20% (i.e., `HR80_20`). We can also generate 95% confidence intervals using bootstrapping.

```{r}
HR80_20 = get_hr(test_data, CI = TRUE, boot = 300)
str(HR80_20)
```

Each `get_` function also returns the output from each bootstrap iteration in `$iters` so that the user can plot these or calculate their own confidence intervals:

```{r HR_histogram}
ggplot(mapping = aes(x = HR80_20$iters)) +
    geom_histogram(binwidth = 2/3) +
    geom_vline(xintercept = HR80_20$HR, color = "red") +
    geom_vline(xintercept = HR80_20$conf.low) +
    geom_vline(xintercept = HR80_20$conf.high) +
        theme_minimal() +
        labs(x = "HR80_20", y = "Count")
```

Similarly, calculate the odds ratio at age 70 between the top 20% and bottom 20% of PHSes.

```{r}
OR80_20 = get_or(test_data, or_age = 70, CI = TRUE, boot = 300)
str(OR80_20)
```

```{r OR_histogram}
ggplot(mapping = aes(x = OR80_20$iters)) +
    geom_histogram(binwidth = 3/2) +
    geom_vline(xintercept = OR80_20$OR, color = "red") +
    geom_vline(xintercept = OR80_20$conf.low) +
    geom_vline(xintercept = OR80_20$conf.high) +
        theme_minimal() +
        labs(x = "OR80_20", y = "Count")
```

Return the concordance index with 95% confidence intervals from a coxph fit:

```{r}
c_index = get_cindex(test_data, CI = TRUE, boot = 300)
str(c_index)
```

Finally, plot the Kaplan-Meier curves with confidence intervals for centiles of interest.

```{r plot_km_curves, fig.width=4.5}
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
    x = x * 100
    y = y * 100
    out = paste0("PHS ", x, "-", y, "th centile")
    return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
    curven <- km_curve(data = test_data,  
                       interval = curves[[i]],
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
    curven$label = label_generator(curves[1, i], curves[2, i])
    km_curves = rbind(km_curves, curven)
}


ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
    geom_ribbon(alpha = 0.1,
                color = 0) +
    geom_step() +
    theme_minimal() +
    xlim(40, 100) + 
    ylim(0, 1) +
    labs(x = "Age", y = "Disease-free Survival") +
    scale_color_brewer(palette = "Set1",
                       name = "Centile") +
    scale_fill_brewer(palette = "Set1",
                       name = "Centile")
```

## More advanced topics

### Data format

Each of the functions in this package (e.g., `get_hr`, `get_or`, `phs_hist`, etc.)  can accept data in several formats. If your `data.frame` contains standard columns `phs`, `age`, and `status`, then you can simply pass the data.frame object to the `data = ` argument:

```{r}
get_hr(data = test_data) |> 
    str()
```

If your data is stored in a data.frame with non-standard column names, e.g., `HazardScores`, `CaseControl`, and `AgeLastObs`,

```{r}
test_data2 = with(test_data, data.frame(HazardScores = phs, 
                                        CaseControl = status, 
                                        AgeLastObs = age))
head(test_data2)
```

then you can specify these column names as separate arguments:

```{r}
get_hr(data = test_data2, 
       phs = "HazardScores", 
       status = "CaseControl", 
       age = "AgeLastObs") |> 
    str()
```

If any of your data is not in a data.frame, but is instead in separate vectors, you can provide these individually by name:

```{r}
hazScore = test_data$phs
caseCon = test_data$status
ageLastObs = test_data$age

get_hr(phs = hazScore, 
       status = caseCon, 
       age = ageLastObs) |> 
    str()
```

Lastly, you can do any combination of the above:

```{r}
test_data3 = data.frame(phs = test_data$phs,
                        CaseControl = test_data$status)
Ages = test_data$age

get_hr(data = test_data3,
       #phs = "phs", # omitted because test_data3 already contains column "phs"
       status = "CaseControl", 
       age = Ages) |> 
    str()
```

### Hazard ratios and odds ratios

#### Specifying intervals

The `get_hr` function allows several options for customizing your hazard ratios. By default, it calculates the `HR80_20`, which compares the top 20% of PHSes to the bottom 20%. However, custom intervals can be specified using the `upper_interval` and `lower_interval` arguments.

```{r}
get_hr(test_data,
       lower_interval = 0.2,
       upper_interval = 0.8) |> 
    str()
```

These are the default arguments. If one value is passed to `upper_interval` then it calculates the hazard ratio for the interval `0.8:Inf`. Similarly, `lower_interval` will use `-Inf:0.2`. However, two values can be specified as a vector to specify a particular range. For example, if you want the lower interval to be the middle 40% of PHSes, you can specify `lower_interval = c(0.3, 0.7)`. 

```{r}
get_hr(test_data,
       lower_interval = c(0.3, 0.7),
       upper_interval = c(0.8, Inf)) |> 
    str()
```

interval specifications are the same for odds ratios from `get_or` as for `get_hr`. The only difference is that the odds ratio must be calculated at a particular age, which is passed with the `or_age` argument:

```{r}
get_or(test_data,
       or_age = 70,
       lower_interval = c(0.3, 0.7),
       upper_interval = c(0.8, Inf)) |> 
    str()
```

#### Confidence intervals from bootstrapping

95% confidence intervals for the estimates from the `get_hr`, `get_or`,  and `get_cindex` functions can be generated by bootstrapping. This is done by by specifying `CI = TRUE` and providing a number of bootstrap iterations with `bootstrap_iterations = `. In this case, `get_hr` returns a list containing the hazard ratio and the values for the upper and lower 95% confidence intervals as well as all of the HR estimates from each bootstrap iteration in `iters`.

```{r}
get_hr(test_data,
       CI = TRUE,
       bootstrap_iterations = 300) |> 
    str()
```

#### Sample weight correction

This is an experimental feature and hasn't been tested. If you wish to perform sample weight correction for HR, OR or C-Index, you can enable it with the flag `swc = TRUE`. You must provide the population number of cases and the population number of controls to perform sample weight correction. These are provided with the `swc_popnumcases` and `swc_popnumcontrols` arguments.

```{r}
get_hr(test_data,
       swc = TRUE,
       swc_popnumcases = 28421,
       swc_popnumcontrols = 398475) |> 
    str()
```


### Plotting curves

To plot Kaplan-Meier curves, you can use the `km_curve` function to return a single Kaplan-Meier curve for a given interval. Plotting must be performed manually once you have returned these curves.

```{r}
curven <- km_curve(data = test_data,  
                       interval = c(0.80, 1.00), 
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
```

`interval` specifies the lower and upper quantile of the interval. In this case, the K-M curve for the top 20% of samples (`0.8:1.0`) will be returned. In addition, `age_range` specifies the ages over which the curve should be generated. Can be on the order of `40:100` for age-at-diagnosis or similar. Alternatively, can be `0:10` for, for exmaple, age of conversion to AD after MCI diagnosis.

We can generate multiple curves at various intervals and get them into a data.frame:

```{r}
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
    x = x * 100
    y = y * 100
    out = paste0("PHS ", x, "-", y, "th centile")
    return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
    curven <- km_curve(data = test_data,  
                       interval = curves[[i]],
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
    curven$label = label_generator(curves[1, i], curves[2, i])
    km_curves = rbind(km_curves, curven)
}
```

And plot them:

```{r plot_km_curves2, fig.width=4.5}
ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
    geom_ribbon(alpha = 0.1,
                color = 0) +
    geom_step() +
    theme_minimal() +
    xlim(40, 100) + 
    ylim(0, 1) +
    labs(x = "Age", y = "Disease-free Survival") +
    scale_color_brewer(palette = "Set1",
                       name = "Centile") +
    scale_fill_brewer(palette = "Set1",
                       name = "Centile")
```


### Individual Prediction

It is possible to generate a lookup table to perform individual prediction on a new, un-observed individual for which we have calculated a PHS score. To do this, we fit the model using our data and generate predicted curves for each of 99 percentiles (1 to 99%). We then output these curves with their associated PHS score at each percentile. This removes any connection to the subject-level data so is safe to export from secure computing platforms. We can then plot the prediction for the closest percentile to the new individual's PHS

Future work will be to implement a way of interpolating the curve for a subject that is between two percentiles. In addition, work should be done to perform smoothing of the curve using the baseline hazard function rather than outputing a K-M curve for that individual. This will improve the interpretability of the individual prediction since risk isn't likely to suddenly drop-off at a particular age, but should be an exponential function.

Here is an example workflow:

```{r create_lookup_table}
# Create the lookup table (on a secure computing platform using
# individual-level data)

lookup_table = create_lookup_table(test_data)
```

Export that lookup table from the computing platform and then identify the percentile for a new subject. Based on this toy dataset, a subject with a PHS of 1.6 would be approximately in the 94th percentile. Here, I plot the predicted curve for the median percentiel (50%) with a gray shaded area for the confidence intervals, an individual line for every percentile from 1 to 99, and I have highlighted in red the predicted risk curve and confidence intervals for the "new" hypothetical subject with a PHS of 1.6 (94th percentile):

```{r plot_ind_pred}
ggplot(lookup_table[lookup_table$percentile == 0.50, ], 
       aes(x = time, y = surv, ymin = lower, ymax = upper)) +
    geom_line() +
    geom_ribbon(alpha = 0.5) +
    theme_classic() +
    geom_line(aes(x = time, y = surv, group = percentile), data = lookup_table, alpha = 0.1) +
    geom_line(aes(x = time, y = surv, group = percentile), 
                data = lookup_table[lookup_table$percentile == 0.94, ], 
                color = "red") +
    geom_ribbon(aes(x = time, y = surv, ymin = lower, ymax = upper), 
                data = lookup_table[lookup_table$percentile == 0.94, ], 
                alpha = 0.5, fill = "red") +
    labs(x = "Age", y = "Disease-free Survival")
```


## Example Script

Here is an example script to run the entire hazrd analysis from the command line:

```
Rscript hazrd_script.R phs.tsv metadata.tsv FALSE reports/
```

```{r, eval=FALSE}
#########################################
# First argument: PHS file
# tab delimited file, with or without header, first column ID, second column PHS score
# file basename used as prefix for output
#
# Second argument: metadata file
# tab delimited file with or without header, first column id,  second column age, 
# third column status (0 or 1). Additional columns will be ignored.
#
# Third argument: 
# boolean (i.e., TRUE or FALSE) indicating whether to inverse (x * -1) the PHS scores
# to reverse the direction of effect.
#
# Fourth argument:
# path to output directory, e.g., "reports/"
# if the directory does not exist it will be created

########################################
# Load packages
library(hazrd)
library(ggplot2)


#######################################
# parse arguments

args = commandArgs(trailingOnly=TRUE)

model_file = args[1] # PHS file

# model = gsub("\\_.*","", basename(args[1])) # filename for output
model = gsub("\\..*","", basename(args[1])) # filename for output

metadata = read.table(args[2], col.names = c("id", "age", "status")) # Phenotype data

inverse = args[3]

output = args[4]

#########################################
# Import PHS data

phs = read.table(model_file, col.names = c("id", "phs"))

print(head(phs))

phs$phs = scale(phs$phs, center = TRUE, scale = TRUE)

if (inverse) {phs$phs = phs$phs * -1}

combined_data = merge(metadata, phs, by = "id")

combined_data = combined_data[!(is.na(combined_data$phs)), ]

#########################################
# Analyze PHS data

# PHS Histogram
phs_hist = 
	phs_hist(combined_data, normalize = TRUE) +
	labs(title = model) + theme(aspect.ratio=1)

ggsave(paste0(output, model, "_hist.png"), phs_hist, 
       units = "px", height = 1300, width = 1300, create.dir = TRUE)

# Performance metrics


HR80_20 = get_hr(combined_data, CI = TRUE, boot = 300)
ages = c(60, 70, 80)
OR80_20 = lapply(ages, function(x) {get_or(combined_data, or_age = x, CI = TRUE, boot = 300)})
c_index = get_cindex(combined_data, CI = TRUE, boot = 300)

HR80_20$age = NA
c_index$age = NA
indices = list(HR80_20, OR80_20[[1]], OR80_20[[2]], OR80_20[[3]], c_index)

perform = do.call(rbind.data.frame, lapply(indices, function(x){data.frame(index = x$index, 
                                                                 value = x$value, 
                                                                 conf.low = x$conf.low, 
                                                                 conf.high = x$conf.high, 
                                                                 age = x$age)}))
perform = cbind(model, perform)
write.table(perform, paste0(output, model, "_performance.tsv"), sep = "\t", row.names = FALSE)

# KM Curves
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
  x = x * 100
  y = y * 100
  out = paste0("PHS ", x, "-", y, "th centile")
  return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
  curven <- km_curve(data = combined_data,  
                     interval = curves[[i]],
                     age_range = 40:100, 
                     scale = FALSE, 
                     inverse = FALSE)
  curven$label = label_generator(curves[1, i], curves[2, i])
  km_curves = rbind(km_curves, curven)
}


km_curve = ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
  geom_ribbon(alpha = 0.1,
              color = 0) +
  geom_step() +
  theme_minimal() +
  xlim(40, 100) + 
  ylim(0, 1) +
  labs(x = "Age", y = "Disease-free Survival", title = model) +
  scale_color_brewer(palette = "Set1",
                     name = "Centile") +
  scale_fill_brewer(palette = "Set1",
                    name = "Centile")  + theme(aspect.ratio=1)

ggsave(paste0(output, model, "_km_curve.png"), km_curve,
       units = "px", height = 1300, width = 1600, create.dir = TRUE)
```

