---
title: "Getting started with hazrd"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with hazrd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hazrd)
library(ggplot2)
```

`hazrd` provides a consistent API for computing survival-based discrimination
metrics and visualizing Kaplan-Meier curves stratified by polygenic hazard score
(PHS). The primary user-facing functions are `phs_metrics()` for statistics,
`phs_km_curve()` for Kaplan-Meier plots, and `phs_cox_curve()` for Cox
model-based survival curves at specified PHS percentiles.

## Data

The package ships with a simulated survival dataset, `test_data`. The three
required columns are `phs` (continuous score), `age` (time to event or
censoring), and `status` (event indicator: 1 = event, 0 = censored).

Here is our test data included in the package.

```{r}
head(test_data)
```

## Computing metrics with `phs_metrics()`

`phs_metrics()` is the single entry point for all discrimination statistics. It
fits a Cox proportional hazards model and derives metrics from the fit. The
`metrics` argument controls which statistics are computed; column names are
passed as strings and default to `"phs"`, `"age"`, and `"status"`.

### All core metrics in one call

```{r all_metrics}
metrics <- phs_metrics(
  test_data,
  metrics  = c("HR", "C_index", "OR", "HR_SD"),
  or_age   = 70
)
metrics
```

The returned tibble has one row per metric. `conf_low`, `conf_high`, and `se`
are `NA` until bootstrapping is enabled.

### Hazard ratio

By default `phs_metrics()` computes `HR[80-100]_[0-20]` — the ratio of the top
20% to the bottom 20% of the PHS distribution. Use `hr_numerator` /
`hr_denominator` to change the bands:

```{r hr_custom}
phs_metrics(test_data, metrics = "HR", hr_numerator = 0.90, hr_denominator = 0.10)
```

For multiple HRs in one call, supply `hr_pairs`:

```{r hr_pairs}
phs_metrics(
  test_data,
  metrics  = "HR",
  hr_pairs = list(
    list(numerator = c(0.80, 1.00), denominator = c(0.00, 0.20)),
    list(numerator = c(0.80, 1.00), denominator = c(0.40, 0.60))
  )
)
```

### Odds ratio

The OR is computed from Kaplan-Meier survival estimates at a specific age.
`or_age` accepts a vector; one row is returned per age:

```{r or}
phs_metrics(test_data, metrics = "OR", or_age = c(65, 70, 75))
```

### Bootstrapped confidence intervals

Set `bootstrap = TRUE` to populate `conf_low`, `conf_high`, and `se`. Use a
small `n_boot` for speed during development:

```{r bootstrap}
phs_metrics(
  test_data,
  metrics   = c("HR", "C_index", "HR_SD"),
  bootstrap = TRUE,
  n_boot    = 300,
  seed      = 42
)
```

### Non-default column names

Column names are passed as strings, making `phs_metrics()` straightforward to
use with non-standard data frames:

```{r nonstandard}
test_data2 <- with(test_data, data.frame(
  score     = phs,
  diagnosis_age = age,
  case      = status
))

phs_metrics(
  test_data2,
  phs   = "score",
  time  = "diagnosis_age",
  event = "case",
  metrics = c("HR", "C_index")
)
```

## Kaplan-Meier curves with `phs_km_curve()`

`phs_km_curve()` stratifies the cohort into percentile bands and plots
empirical Kaplan-Meier survival curves for each band. It returns either a
`ggplot` object (`output = "plot"`, the default) or a tidy data frame
(`output = "data"`).

### Default plot

The default `breaks = c(0.20, 0.80)` splits the cohort into bottom 20%,
middle 60%, and top 20%:

```{r km_default}
phs_km_curve(test_data)
```

### Custom cutpoints

Pass any numeric vector of percentile cutpoints (strictly in (0, 1)) to
`breaks`:

```{r km_custom}
# Quintiles
phs_km_curve(test_data, breaks = c(0.20, 0.40, 0.60, 0.80))
```

### Modifying the ggplot output

`phs_km_curve()` returns a standard `ggplot` object that can be extended with
any ggplot2 layers, themes, or scales:

```{r km_styled}
phs_km_curve(test_data) +
  theme_classic(base_size = 13) +
  labs(
    title = "Disease-free Survival by PHS Percentile",
    x     = "Age (years)",
    y     = "Disease-free Probability",
    color = "PHS Group",
    fill  = "PHS Group"
  ) +
  coord_cartesian(xlim = c(40, 100))
```

### Getting the underlying data

Use `output = "data"` when you want to build a fully custom plot or pass the
survival estimates on to further analysis:

```{r km_data}
km_data <- phs_km_curve(test_data, output = "data")
head(km_data)
```

The returned data frame has columns `time`, `estimate`, `conf.low`, `conf.high`,
`n.risk`, `n.event`, and `stratum`, which can be used directly with ggplot2:

```{r km_manual_plot}
ggplot(km_data, aes(x = time, y = estimate,
                    ymin = conf.low, ymax = conf.high,
                    color = stratum, fill = stratum)) +
  geom_ribbon(alpha = 0.15, color = NA) +
  geom_step(linewidth = 0.8) +
  scale_color_brewer(palette = "Set1", name = "PHS Group") +
  scale_fill_brewer(palette = "Set1", name = "PHS Group") +
  coord_cartesian(xlim = c(40, 100), ylim = c(0, 1)) +
  labs(x = "Age (years)", y = "Disease-free Probability") +
  theme_minimal()
```

## Cox model curves with `phs_cox_curve()`

`phs_cox_curve()` fits a Cox proportional-hazards model (`phs` as the sole
predictor) and returns predicted survival curves for individuals at specified
PHS percentiles. Unlike `phs_km_curve()`, these are smooth model-based
predictions, not empirical group estimates.

```{r cox_default}
phs_cox_curve(test_data)
```

Supply specific percentiles via the `percentiles` argument:

```{r cox_custom}
phs_cox_curve(test_data, percentiles = c(0.01, 0.10, 0.50, 0.90, 0.99))
```
