---
title: "hazrd"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hazrd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hazrd)
library(ggplot2)
set.seed(46495809)
```

## Getting started with `{hazrd}`

First, generate some test data.

```{r}

n = 1000
status = rbinom(n, 1, 0.2)

test_data = data.frame(phs  = rnorm(n) + (1 * status),
                       status = status,
                       age = sample(40:100, n, replace = TRUE))
```

Next, plot the histogram of PHSes by case/control status.

```{r phs_hist, fig.width=4}
phs_hist(test_data, normalize = TRUE)
```

### Generating test statistics and confidence intervals

Then, calculate the hazard ratio comparing the mean of the top 20% of PHSes to the mean of the bottom 20% (i.e., `HR80_20`). We can also generate 95% confidence intervals using bootstrapping.

```{r}
HR80_20 = get_hr(test_data, CI = TRUE, boot = 300)
print(HR80_20)
```

Similarly, calculate the odds ratio at age 70 between the top 20% and bottom 20% of PHSes.

```{r}
OR80_20 = get_or(test_data, or_age = 70, CI = TRUE, boot = 300)
print(OR80_20)
```

Return the concordance index with 95% confidence intervals from a coxph fit:

```{r}
c_index = get_cindex(test_data,  CI = TRUE, boot = 300)
print(c_index)
```

Finally, plot the Kaplan-Meier curves with confidence intervals for centiles of interest.

```{r generate_km_curve, fig.width=4.5}
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
    x = x * 100
    y = y * 100
    out = paste0("PHS ", x, "-", y, "th centile")
    return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
    curven <- km_curve(data = test_data,  
                       interval = curves[[i]],
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
    curven$label = label_generator(curves[1, i], curves[2, i])
    km_curves = rbind(km_curves, curven)
}


ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
    geom_ribbon(alpha = 0.1,
                color = 0) +
    geom_step() +
    theme_minimal() +
    xlim(40, 100) + 
    ylim(0, 1) +
    labs(x = "Age", y = "Disease-free Survival") +
    scale_color_brewer(palette = "Set1",
                       name = "Centile") +
    scale_fill_brewer(palette = "Set1",
                       name = "Centile")
```

## More advanced topics

### Data format

Each of the functions in this package (e.g., `get_hr`, `get_or`, `phs_hist`, etc.)  can accept data in several formats. If your `data.frame` contains standard columns `phs`, `age`, and `status`, then you can simply pass the data.frame object to the `data = ` argument:

```{r}
get_hr(data = test_data)
```

If your data is stored in a data.frame with non-standard column names, e.g., `HazardScores`, `CaseControl`, and `AgeLastObs`,

```{r}
test_data2 = with(test_data, data.frame(HazardScores = phs, 
                                        CaseControl = status, 
                                        AgeLastObs = age))
head(test_data2)
```

then you can specify these column names as separate arguments:

```{r}
get_hr(data = test_data2, 
       phs = "HazardScores", 
       status = "CaseControl", 
       age = "AgeLastObs")
```

If any of your data is not in a data.frame, but is instead in separate vectors, you can provide these individually by name:

```{r}
hazScore = test_data$phs
caseCon = test_data$status
ageLastObs = test_data$age

get_hr(phs = hazScore, 
       status = caseCon, 
       age = ageLastObs)
```

Lastly, you can do any combination of the above:

```{r}
test_data3 = data.frame(phs = test_data$phs,
                        CaseControl = test_data$status)
Ages = test_data$age

get_hr(data = test_data3,
       #phs = "phs", # omitted because test_data3 already contains standard column "phs"
       status = "CaseControl", 
       age = Ages)
```

### Hazard ratios and odds ratios

#### Specifying quantiles

The `get_hr` function allows several options for customizing your hazard ratios. By default, it calculates the `HR80_20`, which compares the top 20% of PHSes to the bottom 20%. However, custom quantiles can be specified using the `upper_quantile` and `lower_quantile` arguments.

```{r}
get_hr(test_data,
       upper_quantile = 0.8,
       lower_quantile = 0.2)
```

These are the default arguments. If one value is passed to `upper_quantile` then it calculates the hazard ratio for the quantile `0.8:Inf`. Similarly, `lower_quantile` will use `-Inf:0.2`. However, two values can be specified as a vector to specify a particular range. For example, if you want the lower quantile to be the middle 40% of PHSes, you can specify `lower_quantile = c(0.3, 0.7)`. 

```{r}
get_hr(test_data,
       upper_quantile = c(0.8, Inf),
       lower_quantile = c(0.3, 0.7))
```

Quantile specifications are the same for odds ratios from `get_or` as for `get_hr`. The only difference is that the odds ratio must be calculated at a particular age, which is passed with the `or_age` argument:

```{r}
get_or(test_data,
       or_age = 70,
       upper_quantile = c(0.8, Inf),
       lower_quantile = c(0.3, 0.7))
```

#### Confidence intervals from bootstrapping

95% confidence intervals for the estimates from the `get_hr`, `get_or`,  and `get_cindex` functions can be generated by bootstrapping. This is done by by specifying `boot = TRUE` and providing a number of bootstrap iterations with `B = `. In this case, instead of a single value, `get_hr` returns a list containing the hazard ratio and the values for the upper and lower 95% confidence intervals.

```{r}
get_hr(test_data,
       boot = TRUE,
       B = 300)
```

#### Sample weight correction

This is an experimental feature and hasn't been tested. If you wish to perform sample weight correction for HR, OR or C-Index, you can enable it with the flag `swc = TRUE`. You must provide the population number of cases and the population number of controls to perform sample weight correction. These are provided with the `swc_popnumcases` and `swc_popnumcontrols` arguments.

```{r}
get_hr(test_data,
       swc = TRUE,
       swc_popnumcases = 28421,
       swc_popnumcontrols = 398475)
```


### Plotting curves

To plot Kaplan-Meier curves, you can use the `km_curve` function to return a single Kaplan-Meier curve for a given quantile group. Plotting must be performed manually once you have returned these curves.

```{r}
curven <- km_curve(data = test_data,  
                       lower = 0.80,
                       upper = 1.00, 
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
```

`lower` specifies the bottom of the quantile and `upper` specifies the top of the quantile. In this case, the K-M curve for the top 20% of samples (`0.8:1.0`) will be returned. In addition, `age_range` specifies the ages over which the curve should be generated. Can be on the order of `40:100` for age-at-diagnosis or similar. Alternatively, can be `0:10` for, for exmaple, age of conversion to AD after MCI diagnosis.