---
title: "Getting started with `{hazrd}`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hazrd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hazrd)
library(ggplot2)
```

Here is our test data included in the package.

```{r}
head(test_data)
```

Next, plot the histogram of PHSes by case/control status.

```{r phs_hist, fig.width=4}
phs_hist(test_data, normalize = TRUE)
```

### Generating test statistics and confidence intervals

Then, calculate the hazard ratio comparing the mean of the top 20% of PHSes to the mean of the bottom 20% (i.e., `HR80_20`). We can also generate 95% confidence intervals using bootstrapping.

```{r}
HR80_20 = get_hr(test_data, CI = TRUE, boot = 300)
str(HR80_20)
```

Each `get_` function also returns the output from each bootstrap iteration in `$iters` so that the user can plot these or calculate their own confidence intervals:

```{r HR_histogram}
ggplot(mapping = aes(x = HR80_20$iters)) +
    geom_histogram(binwidth = 2/3) +
    geom_vline(xintercept = HR80_20$HR, color = "red") +
    geom_vline(xintercept = HR80_20$conf.low) +
    geom_vline(xintercept = HR80_20$conf.high) +
        theme_minimal() +
        labs(x = "HR80_20", y = "Count")
```

Similarly, calculate the odds ratio at age 70 between the top 20% and bottom 20% of PHSes.

```{r}
OR80_20 = get_or(test_data, or_age = 70, CI = TRUE, boot = 300)
str(OR80_20)
```

```{r OR_histogram}
ggplot(mapping = aes(x = OR80_20$iters)) +
    geom_histogram(binwidth = 3/2) +
    geom_vline(xintercept = OR80_20$OR, color = "red") +
    geom_vline(xintercept = OR80_20$conf.low) +
    geom_vline(xintercept = OR80_20$conf.high) +
        theme_minimal() +
        labs(x = "OR80_20", y = "Count")
```

Return the concordance index with 95% confidence intervals from a coxph fit:

```{r}
c_index = get_cindex(test_data, CI = TRUE, boot = 300)
str(c_index)
```

Finally, plot the Kaplan-Meier curves with confidence intervals for centiles of interest.

```{r plot_km_curves}
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
    x = x * 100
    y = y * 100
    out = paste0("PHS ", x, "-", y, "th centile")
    return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
    curven <- km_curve(data = test_data,  
                       interval = curves[[i]],
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
    curven$label = label_generator(curves[1, i], curves[2, i])
    km_curves = rbind(km_curves, curven)
}


ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
    geom_ribbon(alpha = 0.1,
                color = 0) +
    geom_step() +
    theme_minimal() +
    xlim(40, 100) + 
    ylim(0, 1) +
    labs(x = "Age", y = "Disease-free Survival") +
    scale_color_brewer(palette = "Set1",
                       name = "Centile") +
    scale_fill_brewer(palette = "Set1",
                       name = "Centile")
```

## More advanced topics

### Data format

Each of the functions in this package (e.g., `get_hr`, `get_or`, `phs_hist`, etc.)  can accept data in several formats. If your `data.frame` contains standard columns `phs`, `age`, and `status`, then you can simply pass the data.frame object to the `data = ` argument:

```{r}
get_hr(data = test_data) |> 
    str()
```

If your data is stored in a data.frame with non-standard column names, e.g., `HazardScores`, `CaseControl`, and `AgeLastObs`,

```{r}
test_data2 = with(test_data, data.frame(HazardScores = phs, 
                                        CaseControl = status, 
                                        AgeLastObs = age))
head(test_data2)
```

then you can specify these column names as separate arguments:

```{r}
get_hr(data = test_data2, 
       phs = "HazardScores", 
       status = "CaseControl", 
       age = "AgeLastObs") |> 
    str()
```

If any of your data is not in a data.frame, but is instead in separate vectors, you can provide these individually by name:

```{r}
hazScore = test_data$phs
caseCon = test_data$status
ageLastObs = test_data$age

get_hr(phs = hazScore, 
       status = caseCon, 
       age = ageLastObs) |> 
    str()
```

Lastly, you can do any combination of the above:

```{r}
test_data3 = data.frame(phs = test_data$phs,
                        CaseControl = test_data$status)
Ages = test_data$age

get_hr(data = test_data3,
       #phs = "phs", # omitted because test_data3 already contains column "phs"
       status = "CaseControl", 
       age = Ages) |> 
    str()
```

### Hazard ratios and odds ratios

#### Specifying intervals

The `get_hr` function allows several options for customizing your hazard ratios. By default, it calculates the `HR80_20`, which compares the top 20% of PHSes to the bottom 20%. However, custom intervals can be specified using the `upper_interval` and `lower_interval` arguments.

```{r}
get_hr(test_data,
       lower_interval = 0.2,
       upper_interval = 0.8) |> 
    str()
```

These are the default arguments. If one value is passed to `upper_interval` then it calculates the hazard ratio for the interval `0.8:Inf`. Similarly, `lower_interval` will use `-Inf:0.2`. However, two values can be specified as a vector to specify a particular range. For example, if you want the lower interval to be the middle 40% of PHSes, you can specify `lower_interval = c(0.3, 0.7)`. 

```{r}
get_hr(test_data,
       lower_interval = c(0.3, 0.7),
       upper_interval = c(0.8, Inf)) |> 
    str()
```

interval specifications are the same for odds ratios from `get_or` as for `get_hr`. The only difference is that the odds ratio must be calculated at a particular age, which is passed with the `or_age` argument:

```{r}
get_or(test_data,
       or_age = 70,
       lower_interval = c(0.3, 0.7),
       upper_interval = c(0.8, Inf)) |> 
    str()
```

#### Confidence intervals from bootstrapping

95% confidence intervals for the estimates from the `get_hr`, `get_or`,  and `get_cindex` functions can be generated by bootstrapping. This is done by by specifying `CI = TRUE` and providing a number of bootstrap iterations with `bootstrap_iterations = `. In this case, `get_hr` returns a list containing the hazard ratio and the values for the upper and lower 95% confidence intervals as well as all of the HR estimates from each bootstrap iteration in `iters`.

```{r}
get_hr(test_data,
       CI = TRUE,
       bootstrap_iterations = 300) |> 
    str()
```

#### Sample weight correction

This is an experimental feature and hasn't been tested. If you wish to perform sample weight correction for HR, OR or C-Index, you can enable it with the flag `swc = TRUE`. You must provide the population number of cases and the population number of controls to perform sample weight correction. These are provided with the `swc_popnumcases` and `swc_popnumcontrols` arguments.

```{r}
get_hr(test_data,
       swc = TRUE,
       swc_popnumcases = 28421,
       swc_popnumcontrols = 398475) |> 
    str()
```


### Plotting curves

To plot Kaplan-Meier curves, you can use the `km_curve` function to return a single Kaplan-Meier curve for a given interval. Plotting must be performed manually once you have returned these curves.

```{r}
curven <- km_curve(data = test_data,  
                       interval = c(0.80, 1.00), 
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
```

`interval` specifies the lower and upper quantile of the interval. In this case, the K-M curve for the top 20% of samples (`0.8:1.0`) will be returned. In addition, `age_range` specifies the ages over which the curve should be generated. Can be on the order of `40:100` for age-at-diagnosis or similar. Alternatively, can be `0:10` for, for exmaple, age of conversion to AD after MCI diagnosis.

We can generate multiple curves at various intervals and get them into a data.frame:

```{r}
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
    x = x * 100
    y = y * 100
    out = paste0("PHS ", x, "-", y, "th centile")
    return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
    curven <- km_curve(data = test_data,  
                       interval = curves[[i]],
                       age_range = 40:100, 
                       scale = FALSE, 
                       inverse = FALSE)
    curven$label = label_generator(curves[1, i], curves[2, i])
    km_curves = rbind(km_curves, curven)
}
```

And plot them:

```{r plot_km_curves2}
ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
    geom_ribbon(alpha = 0.1,
                color = 0) +
    geom_step() +
    theme_minimal() +
    xlim(40, 100) + 
    ylim(0, 1) +
    labs(x = "Age", y = "Disease-free Survival") +
    scale_color_brewer(palette = "Set1",
                       name = "Centile") +
    scale_fill_brewer(palette = "Set1",
                       name = "Centile")
```


### Individual Prediction

It is possible to generate a lookup table to perform individual prediction on a new, un-observed individual for which we have calculated a PHS score. To do this, we fit the model using our data and generate predicted curves for each of 99 percentiles (1 to 99%). We then output these curves with their associated PHS score at each percentile. This removes any connection to the subject-level data so is safe to export from secure computing platforms. We can then plot the prediction for the closest percentile to the new individual's PHS

Here is an example workflow:

```{r create_lookup_table}
# Create the lookup table (on a secure computing platform using
# individual-level data)

lookup_table = create_lookup_table(test_data)
```

Export that lookup table from the computing platform and then identify the percentile for a new subject. Based on this toy dataset, a subject with a PHS of 1.6 would be approximately in the 94th percentile. Here, I plot the predicted curve for the median percentiel (50%) with a gray shaded area for the confidence intervals, an individual line for every percentile from 1 to 99, and I have highlighted in red the predicted risk curve and confidence intervals for the "new" hypothetical subject with a PHS of 1.6 (94th percentile):

```{r plot_ind_pred}
ggplot(lookup_table[lookup_table$percentile == 0.50, ], 
       aes(x = time, y = surv, ymin = lower, ymax = upper)) +
    geom_line() +
    geom_ribbon(alpha = 0.5) +
    theme_classic() +
    geom_line(aes(x = time, y = surv, group = percentile), data = lookup_table, alpha = 0.1) +
    geom_line(aes(x = time, y = surv, group = percentile), 
                data = lookup_table[lookup_table$percentile == 0.94, ], 
                color = "red") +
    geom_ribbon(aes(x = time, y = surv, ymin = lower, ymax = upper), 
                data = lookup_table[lookup_table$percentile == 0.94, ], 
                alpha = 0.5, fill = "red") +
    labs(x = "Age", y = "Disease-free Survival")
```


## Percentiles versus Quintiles for comparison and prediction

In the literature on PHS, there are at least two ways that folks visualize the survival estimates for different PHS models. One is to compare the 20th percentile to the 80th percentile. The other is to compare quintiles, that is, the top 20% of PHSes to the bottom 20%. Here is an explanation of how to do both and how those estimates compare:

### Percentiles

```{r}
library(survival)
# Fit the coxph model
cp_fit = coxph(Surv(age, status) ~ phs, data = test_data)

# Calculate the PHS at each percentile of interest
phs0.2 = as.numeric(quantile(test_data$phs, 0.2))
phs0.8 = as.numeric(quantile(test_data$phs, 0.8))
phs0.5 = as.numeric(quantile(test_data$phs, 0.5))


# Generate some new data for prediction using these PHS scores
new_data0.2 <- data.frame(status = 1,
                         age = 40:100,
                         phs = phs0.2)
new_data0.8 = new_data0.2
new_data0.8$phs = phs0.8
new_data0.5 = new_data0.2
new_data0.5$phs = phs0.5

# Using our original model, generate a predicted survival curve
cp_predict_0.2 <- predict(cp_fit, new_data0.2, type = "survival")
cp_predict_0.8 <- predict(cp_fit, new_data0.8, type = "survival")
cp_predict_0.5 <- predict(cp_fit, new_data0.5, type = "survival")

ggplot() +
    geom_step(aes(x = age, y = cp_predict_0.2), data = new_data0.2, color = 4) +
    geom_step(aes(x = age, y = cp_predict_0.8), data = new_data0.8, color = 2) +
    geom_step(aes(x = age, y = cp_predict_0.5), data = new_data0.5, color = 3) +
    labs(x = "Age", y = "Disease-free Survival") +
    theme_minimal()
```

Here, the blue line is the 20th percentile, the green line is the 50th percentile, and the red line is the 80th percentile.

### Quintiles

```{r}
# Cut the PHSes into groups based on percentile.
# Here, we are splitting them into 0-20%, 20-80%, and 80-100%
test_data$phs_group <- cut(test_data$phs, c(-Inf, phs0.2, phs0.8, Inf),
                    labels = c("0-20", "20-80", "80-100"))

# Fit the model, this time with group as the predictor rather than the PHS
cp_fit_group <- coxph(Surv(age, status) ~ phs_group, test_data)

# Generate some new data for prediction
new_data0020 <- data.frame(status = 1,
                           age = 40:100,
                           phs_group = "0-20")
new_data2080 <- new_data0020
new_data2080$phs_group = "20-80"
new_data80100 <- new_data0020
new_data80100$phs_group = "80-100"

# Generate a predicted survival curve
cp_predict_0020 <- predict(cp_fit_group, new_data0020, type = "survival")
cp_predict_2080 <- predict(cp_fit_group, new_data2080, type = "survival")
cp_predict_80100 <- predict(cp_fit_group, new_data80100, type = "survival")

# Plot the curves for each group
ggplot() +
    geom_step(aes(x = age, y = cp_predict_0020), data = new_data0020, color = 4) +
    geom_step(aes(x = age, y = cp_predict_80100), data = new_data80100, color = 2) +
    geom_step(aes(x = age, y = cp_predict_2080), data = new_data2080, color = 3) +
    labs(x = "Age", y = "Disease-free Survival") +
    theme_minimal()
```

Here, the blue line is the survival curve for the bottom 20% of PHSes, the green line is the survival curve for the middle 60% of PHSes (20-80%), and the red line is the survival curve for the top 20% of PHSes.

Lastly, we can compare these two approaches and see how they generate different survival curves:

```{r}
ggplot() +
    geom_step(aes(x = age, y = cp_predict_0020), data = new_data0020, color = 4) +
    geom_step(aes(x = age, y = cp_predict_2080), data = new_data2080, color = 4) +
    geom_step(aes(x = age, y = cp_predict_80100), data = new_data80100, color = 4) +
    geom_step(aes(x = age, y = cp_predict_0.2), data = new_data0.2, color = 2) +
    geom_step(aes(x = age, y = cp_predict_0.8), data = new_data0.8, color = 2) +
    geom_step(aes(x = age, y = cp_predict_0.5), data = new_data0.5, color = 2) +
    labs(x = "Age", y = "Disease-free Survival") +
    theme_minimal()
```

Here, the blue curves are the quintile curves or group-based curves and the red curves are the percentile-based curves.

The 0-20 and 80-100 curves are close to but not exactly the same as the 10% percentile and the 90% percentile:

```{r}
phs0.1 = as.numeric(quantile(test_data$phs, 0.1))
phs0.9 = as.numeric(quantile(test_data$phs, 0.9))
new_data0.1 = new_data0.2
new_data0.1$phs = phs0.1
new_data0.9 = new_data0.2
new_data0.9$phs = phs0.9
cp_predict_0.1 <- predict(cp_fit, new_data0.1, type = "survival")
cp_predict_0.9 <- predict(cp_fit, new_data0.9, type = "survival")

ggplot() +
    geom_step(aes(x = age, y = cp_predict_0020), data = new_data0020, color = 4) +
    geom_step(aes(x = age, y = cp_predict_2080), data = new_data2080, color = 4) +
    geom_step(aes(x = age, y = cp_predict_80100), data = new_data80100, color = 4) +
    labs(x = "Age", y = "Disease-free Survival") +
    theme_minimal() +
    geom_step(aes(x = age, y = cp_predict_0.1), data = new_data0.1, color = 6)  +
    geom_step(aes(x = age, y = cp_predict_0.9), data = new_data0.9, color = 6) 

```

Here again the blue curves are the group-based curves (0-20, 20-80, and 80-100%) and the pink curves are the prediction for the 10th percentile and the 90th percentile.

---

Now, comparing the 90th percentile individual to the "empirical" curve for the subjects in the top quintile:

```{r}
# This is the curve for the 90% percentile
cp_predict_0.9

# This is the curve for the group in the 80-100 quintile
test_data80_100 = test_data[test_data$phs_group == "80-100", ]

surv_80_100 = survfit(Surv(age, status) ~ 1, test_data80_100)
surv_80_100_data = data.frame(age = surv_80_100$time,
                              surv = surv_80_100$surv)

ggplot() + 
    geom_step(aes(x = 40:100, y = cp_predict_0.9), color = 4) + # blue
    geom_step(aes(x = age, y = cp_predict_80100), data = new_data80100, color = 2) + #red
    geom_step(aes(x = age, y = surv), data = surv_80_100_data, color = 3) # green

```

Here, blue is the prediction for the 90th percentile in the full PHS model. Red is the prediction for the 80-100% group from the "PHS Group" model. And green is the "empirical" survival curve for just the individuals in the 80-100% group.

One thing I want to add is that the way to test the proportional hazard assumptions for a model is with the `cox.zph` function.

```{r}
ph_fit = cox.zph(cp_fit)

# Not sig. so proportional hazards assumption is met
print(ph_fit)

# Residual plot visualizing this:
plot(ph_fit)

```



## Example Script

Here is an example script to run the entire hazrd analysis from the command line:

```
Rscript hazrd_script.R phs.tsv metadata.tsv FALSE reports/
```

```{r, eval=FALSE}
#########################################
# First argument: PHS file
# tab delimited file, with or without header, first column ID, second column PHS score
# file basename used as prefix for output
#
# Second argument: metadata file
# tab delimited file with or without header, first column id,  second column age, 
# third column status (0 or 1). Additional columns will be ignored.
#
# Third argument: 
# boolean (i.e., TRUE or FALSE) indicating whether to inverse (x * -1) the PHS scores
# to reverse the direction of effect.
#
# Fourth argument:
# path to output directory, e.g., "reports/"
# if the directory does not exist it will be created

########################################
# Load packages
library(hazrd)
library(ggplot2)


#######################################
# parse arguments

args = commandArgs(trailingOnly=TRUE)

model_file = args[1] # PHS file

# model = gsub("\\_.*","", basename(args[1])) # filename for output
model = gsub("\\..*","", basename(args[1])) # filename for output

metadata = read.table(args[2], col.names = c("id", "age", "status")) # Phenotype data

inverse = args[3]

output = args[4]

#########################################
# Import PHS data

phs = read.table(model_file, col.names = c("id", "phs"))

print(head(phs))

phs$phs = scale(phs$phs, center = TRUE, scale = TRUE)

if (inverse) {phs$phs = phs$phs * -1}

combined_data = merge(metadata, phs, by = "id")

combined_data = combined_data[!(is.na(combined_data$phs)), ]

#########################################
# Analyze PHS data

# PHS Histogram
phs_hist = 
	phs_hist(combined_data, normalize = TRUE) +
	labs(title = model) + theme(aspect.ratio=1)

ggsave(paste0(output, model, "_hist.png"), phs_hist, 
       units = "px", height = 1300, width = 1300, create.dir = TRUE)

# Performance metrics


HR80_20 = get_hr(combined_data, CI = TRUE, boot = 300)
ages = c(60, 70, 80)
OR80_20 = lapply(ages, function(x) {get_or(combined_data, or_age = x, CI = TRUE, boot = 300)})
c_index = get_cindex(combined_data, CI = TRUE, boot = 300)

HR80_20$age = NA
c_index$age = NA
indices = list(HR80_20, OR80_20[[1]], OR80_20[[2]], OR80_20[[3]], c_index)

perform = do.call(rbind.data.frame, lapply(indices, function(x){data.frame(index = x$index, 
                                                                 value = x$value, 
                                                                 conf.low = x$conf.low, 
                                                                 conf.high = x$conf.high, 
                                                                 age = x$age)}))
perform = cbind(model, perform)
write.table(perform, paste0(output, model, "_performance.tsv"), sep = "\t", row.names = FALSE)

# KM Curves
curves = data.frame(curve0_20 = c(0,   0.2),
                    curve20_70 = c(0.2, 0.7),
                    curve80_98 = c(0.8, 0.98),
                    curve98_100 = c(0.98, 1.0))

label_generator = function(x, y) {
  x = x * 100
  y = y * 100
  out = paste0("PHS ", x, "-", y, "th centile")
  return(out)
}

km_curves = data.frame()
for (i in seq_len(length(curves))) {
  curven <- km_curve(data = combined_data,  
                     interval = curves[[i]],
                     age_range = 40:100, 
                     scale = FALSE, 
                     inverse = FALSE)
  curven$label = label_generator(curves[1, i], curves[2, i])
  km_curves = rbind(km_curves, curven)
}


km_curve = ggplot(km_curves, aes(x = time, 
                      y = estimate,
                      ymin = conf.low,
                      ymax = conf.high,
                      col = label,
                      fill = label)) +
  geom_ribbon(alpha = 0.1,
              color = 0) +
  geom_step() +
  theme_minimal() +
  xlim(40, 100) + 
  ylim(0, 1) +
  labs(x = "Age", y = "Disease-free Survival", title = model) +
  scale_color_brewer(palette = "Set1",
                     name = "Centile") +
  scale_fill_brewer(palette = "Set1",
                    name = "Centile")  + theme(aspect.ratio=1)

ggsave(paste0(output, model, "_km_curve.png"), km_curve,
       units = "px", height = 1300, width = 1600, create.dir = TRUE)
```

