---
title: "Getting started with hazrd"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with hazrd}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hazrd)
library(ggplot2)
```

`hazrd` provides a consistent API for computing survival-based discrimination
metrics and visualizing Kaplan-Meier and cumulative incidence curves stratified
by polygenic hazard score (PHS). The primary user-facing functions are
`phs_metrics()` for statistics, `phs_km_curve()` for Kaplan-Meier plots,
`phs_cox_curve()` for Cox model-based survival curves at specified PHS
percentiles, and `phs_abs_risk()` for cumulative incidence (absolute risk)
curves.

## Data

The package ships with a simulated survival dataset, `test_data`. The three
required columns are `phs` (continuous score), `age` (time to event or
censoring), and `status` (event indicator: 1 = event, 0 = censored).

Here is our test data included in the package.

```{r}
head(test_data)
```

## Computing metrics with `phs_metrics()`

`phs_metrics()` is the single entry point for all discrimination statistics. It
fits a Cox proportional hazards model and derives metrics from the fit. The
`metrics` argument controls which statistics are computed; column names are
passed as strings and default to `"phs"`, `"age"`, and `"status"`.

### All core metrics in one call

```{r all_metrics}
metrics <- phs_metrics(
  test_data,
  metrics  = c("HR", "C_index", "OR", "HR_SD"),
  or_age   = 70
)
metrics
```

The returned tibble has one row per metric. `conf_low`, `conf_high`, and `se`
are `NA` until bootstrapping is enabled.

### Hazard ratio

By default `phs_metrics()` computes `HR[80-100]_[0-20]` — the ratio of the top
20% to the bottom 20% of the PHS distribution. Use `hr_numerator` /
`hr_denominator` to change the bands:

```{r hr_custom}
phs_metrics(test_data, metrics = "HR", hr_numerator = 0.90, hr_denominator = 0.10)
```

For multiple HRs in one call, supply `hr_pairs`:

```{r hr_pairs}
phs_metrics(
  test_data,
  metrics  = "HR",
  hr_pairs = list(
    list(numerator = c(0.80, 1.00), denominator = c(0.00, 0.20)),
    list(numerator = c(0.80, 1.00), denominator = c(0.40, 0.60))
  )
)
```

### Odds ratio

The OR is computed from Kaplan-Meier survival estimates at a specific age.
`or_age` accepts a vector; one row is returned per age:

```{r or}
phs_metrics(test_data, metrics = "OR", or_age = c(65, 70, 75))
```

### Bootstrapped confidence intervals

Set `bootstrap = TRUE` to populate `conf_low`, `conf_high`, and `se`. Use a
small `n_boot` for speed during development:

```{r bootstrap}
phs_metrics(
  test_data,
  metrics   = c("HR", "C_index", "HR_SD"),
  bootstrap = TRUE,
  n_boot    = 300,
  seed      = 42
)
```

### Non-default column names

Column names are passed as strings, making `phs_metrics()` straightforward to
use with non-standard data frames:

```{r nonstandard}
test_data2 <- with(test_data, data.frame(
  score     = phs,
  diagnosis_age = age,
  case      = status
))

phs_metrics(
  test_data2,
  phs   = "score",
  time  = "diagnosis_age",
  event = "case",
  metrics = c("HR", "C_index")
)
```

## Kaplan-Meier curves with `phs_km_curve()`

`phs_km_curve()` stratifies the cohort into percentile bands and plots
empirical Kaplan-Meier survival curves for each band. It returns either a
`ggplot` object (`output = "plot"`, the default) or a tidy data frame
(`output = "data"`).

### Default plot

The default `intervals` produce four bands — top 5% (0.95–1), top 20%
(0.80–1), middle 40% (0.30–0.70), and bottom 20% (0–0.20). Bands may overlap,
which is useful for showing how progressively broader high-risk groups compare
to the low-risk reference:

```{r km_default}
phs_km_curve(test_data)
```

### Custom overlapping intervals

Pass a list of `c(lo, hi)` percentile pairs to `intervals`. Bands can overlap
freely — each is an independent Kaplan-Meier fit on the corresponding subset:

```{r km_custom}
phs_km_curve(
  test_data,
  intervals = list(c(0.95, 1), c(0.80, 1), c(0.00, 0.20))
)
```

### Exclusive bands (legacy `breaks`)

For non-overlapping bands, use `breaks` with `intervals = NULL`. Pass a numeric
vector of percentile cutpoints (strictly between 0 and 1):

```{r km_breaks}
# Quintiles
phs_km_curve(test_data, intervals = NULL, breaks = c(0.20, 0.40, 0.60, 0.80))
```

### Modifying the ggplot output

`phs_km_curve()` returns a standard `ggplot` object that can be extended with
any ggplot2 layers, themes, or scales:

```{r km_styled}
phs_km_curve(test_data) +
  theme_classic(base_size = 13) +
  labs(
    title = "Disease-free Survival by PHS Percentile",
    x     = "Age (years)",
    y     = "Disease-free Probability",
    color = "PHS Group",
    fill  = "PHS Group"
  ) +
  coord_cartesian(xlim = c(40, 100))
```

### Getting the underlying data

Use `output = "data"` when you want to build a fully custom plot or pass the
survival estimates on to further analysis:

```{r km_data}
km_data <- phs_km_curve(test_data, output = "data")
head(km_data)
```

The returned data frame has columns `time`, `estimate`, `conf.low`, `conf.high`,
`n.risk`, `n.event`, and `stratum`, which can be used directly with ggplot2:

```{r km_manual_plot}
ggplot(km_data, aes(x = time, y = estimate,
                    ymin = conf.low, ymax = conf.high,
                    color = stratum, fill = stratum)) +
  geom_ribbon(alpha = 0.15, color = NA) +
  geom_step(linewidth = 0.8) +
  scale_color_brewer(palette = "Set1", name = "PHS Group") +
  scale_fill_brewer(palette = "Set1", name = "PHS Group") +
  coord_cartesian(xlim = c(40, 100), ylim = c(0, 1)) +
  labs(x = "Age (years)", y = "Disease-free Probability") +
  theme_minimal()
```

## Cox model curves with `phs_cox_curve()`

`phs_cox_curve()` fits a Cox proportional-hazards model (`phs` as the sole
predictor) and returns predicted survival curves for individuals at specified
PHS percentiles. Unlike `phs_km_curve()`, these are smooth model-based
predictions, not empirical group estimates.

```{r cox_default}
phs_cox_curve(test_data)
```

Supply specific percentiles via the `percentiles` argument:

```{r cox_custom}
phs_cox_curve(test_data, percentiles = c(0.01, 0.10, 0.50, 0.90, 0.99))
```

## Cumulative incidence curves with `phs_abs_risk()`

`phs_abs_risk()` computes absolute (cumulative incidence) risk curves for
percentile-stratified groups. The default `method = "km"` plots 1 minus the
Kaplan-Meier survival estimate on a percentage scale. When competing risks are
present, `method = "aalen_johansen"` uses the Aalen-Johansen estimator.

The `intervals` argument works exactly like in `phs_km_curve()` — the default
four bands (top 5%, top 20%, middle 40%, bottom 20%) are used when `intervals`
is omitted:

```{r abs_default}
phs_abs_risk(test_data)
```

### Custom intervals

```{r abs_custom}
phs_abs_risk(
  test_data,
  intervals = list(c(0.80, 1), c(0.00, 0.20))
)
```

### Getting estimates at specific time points

Use `output = "data"` with `time_points` to extract a tidy table of estimates
at specific ages — useful for reporting absolute risk at clinically meaningful
time points:

```{r abs_timepoints}
phs_abs_risk(test_data, time_points = c(60, 70, 80), output = "data")
```

### Risk table

Use `phs_risk_table()` to build a numbers-at-risk strip as a standalone ggplot,
then combine it with your curve plot using `attach_risk_table()` or patchwork
directly. This keeps both objects as editable ggplots so you can customise them
independently before combining.

```{r abs_risktable, eval=requireNamespace("patchwork", quietly = TRUE)}
# Quick usage: let attach_risk_table() build the table from the data
p <- phs_abs_risk(test_data)
attach_risk_table(p, test_data)
```

For publication-ready plots, build and customise each object separately, then
combine. The `time_points` argument controls at which ages columns appear:

```{r abs_risktable_custom, eval=requireNamespace("patchwork", quietly = TRUE)}
library(patchwork)

p  <- phs_abs_risk(test_data) + ggplot2::theme_classic()
rt <- phs_risk_table(test_data, time_points = c(55, 65, 75, 85))

p / rt + plot_layout(heights = c(3, 1))
```
